Quantifying the locality of operators that translate Stokes $Q$ \& $U$ to scalars $E$ \& $B$

### 2017-10-31 articles/formalism/sections/introduction.tex
resolved the clearpage issue which was occuring with the use of include command. Completed another pass at the non-locality of the E and B kernels

\section{Introduction}

In this work we follow the convection in which bar-ed variables correspond to those in real space, while the tilde-ed variables correspond to those in harmonic space \cite{Zaldarriaga2001a}.

This paper is organized in the following manner: In \sec{sec:pol-intro} we present a primer on the description of CMB polarization on the sphere and introduce the matrix notation which provides a more concise description of the same. In \sec{sec:real_space_operators} we introduce the necessary tools  and discuss the derivations of the real space operators. In \sec{sec:visualize_operator} we evaluate the real space operators and present visualizations of these functions. Here we also discuss the locality of the real space E \& B operators. In \sec{sec:numerical_implementation} we implement these operators to evaluate E \& B  maps from the Stokes parameters Q \& U and compare these maps and their spectra from those derived using Healpix. We conclude with a discussion and the scope of this new method of analyzing CMB polarization in \sec{sec:discussion}.\\ \\ \\

\textbf{Some comments, questions and ideas that occurred to me while writing the draft:}
\begin{enumerate}
\item Using these local convolution can one now understand the E to B leakage occurring due to masking ?

\item Healpix has rings at constant latitude. So the real space convolution is only perfect for pixels near the pole ? What happens when you consider pixels around any arbitrary pixel on a Healpix map, they are probably not arranged in rings. Maybe for this reason working with upgraded map works and finer pixels allow for a better approximation of rings around arbitrary pixels ??
\end{enumerate}

>>>>

\section{Introduction}
In this work we follow the convection in which bar-ed variables correspond to those in real space, while the tilde-ed variables correspond to those in harmonic space \cite{Zaldarriaga2001a}.

This paper is organized in the following manner: In \sec{sec:pol-intro} we present a primer on the description of CMB polarization on the sphere and introduce the matrix notation which provides a more concise description of the same. In \sec{sec:real_space_operators} we introduce the necessary tools  and discuss the derivations of the real space operators. In \sec{sec:visualize_operator} we evaluate the real space operators and present visualizations of these functions. Here we also discuss the locality of the real space E \& B operators. In \sec{sec:numerical_implementation} we implement these operators to evaluate E \& B  maps from the Stokes parameters Q \& U and compare these maps and their spectra from those derived using Healpix. We conclude with a discussion and the scope of this new method of analyzing CMB polarization in \sec{sec:discussion}.

### 2017-10-31 articles/formalism/sections/kernel_discussion.tex
resolved the clearpage issue which was occuring with the use of include command. Completed another pass at the non-locality of the E and B kernels

The real and imaginary part of the kernel $\mathcal{M}$ are identical irrespective of changes in the galactic latitude and longitude of the central pixel. Note that these functions are not distorted when a part of the domain overlaps with the poles, as can be seen in the first three rows of \fig{fig:vis_kernel}. Both these facts can be associated with the fact that this function does not depend on the Euler angle $\gamma$. From \eq{eq:qu2eb_convolution_explicit} and \eq{eq:eb2qu_convolution_explicit} it is clear that $\mathcal{M}_r$ and $\mathcal{M}_i$ can be interpreted in the following ways,
%\revisit{Recall that the angle $\gamma$ defines the rotation about the final z-axis which co-alligns the planar axes. Since the function $\mathcal{M}$ does not depend of this final orientation, it implies a rotational invariance of the fields constructed by convolving with this kernel.}
%
\begin{subequations}
\beqry
\bmat E= -\mathcal{M}_r \\ B =+\mathcal{M}_i  \emat  \leftarrow \bmat Q=\delta(\hat{n} - \hat{n}_j)\\ U=0 \emat ~~;~~ \bmat E= -\mathcal{M}_i \\ B = - \mathcal{M}_r  \emat  \leftarrow \bmat Q=0 \\ U=\delta(\hat{n} - \hat{n}_j) \emat \,,  \\
\bmat Q= -\mathcal{M}_r \\ U = -\mathcal{M}_i  \emat  \leftarrow \bmat E=\delta(\hat{n} - \hat{n}_j)\\ B=0 \emat ~~;~~ \bmat Q= +\mathcal{M}_i \\ U = - \mathcal{M}_r  \emat  \leftarrow \bmat E=0 \\ B=\delta(\hat{n} - \hat{n}_j) \emat \,.
\eeqry
\end{subequations}
%
The kernels $\mathcal{D}$ \& $\mathcal{I}$ vary significantly as a function of galactic latitude of the central pixel as seen in the last four columns of \fig{fig:vis_kernel}. These kernels show a two fold symmetry in the vicinity of the poles and this arises due to Euler angle $\gamma \approx 0$ here and therefore $e^{i2(\alpha \pm \gamma)} \approx e^{i2\alpha}$. Note that in this region, the azimuthal profile of the real and imaginary part of these kernels  is similar to $\mathcal{M}_r$ and $\mathcal{M}_i$ respectively.  This also explains why the imaginary part of the band limited delta function $\mathcal{I}$ contributes just as much as the real part in these regions. On moving to lower latitudes, $\mathcal{D}$ quickly transitions to having a four fold symmetry while $\mathcal{I}$ transitions to being dominated by the real part and behaves more like the conventional delta function. This transition can be most easily understood in the flat sky limit where $\gamma = -\alpha$ which leads to the resultant 4 fold symmetry seen for $\mathcal{D}$ owing to $e^{i2(\alpha - \gamma)} =e^{i4\alpha}$ and $\mathcal{I}$ being dominated by the real part owing to $e^{i2(\alpha + \gamma)} =1 + i0$. Since the flat sky approximation has most validity in the proximity of the equator these limiting tendencies of the respective kernels are seen in the bottom row of \fig{fig:vis_kernel} which depict the kernels evaluated at the equator while the middle two row depict the kernels evaluated at a latitudes of $b=87^{\circ}~\&~ 80^{\circ}$ serve to indicate the rate of this transition. These kernels are invariant under changes in longitude of the central pixel, the latitude being held fixed as one may have expected.

\subsection{Quantifying the non-locality of E \& B modes} \label{sec:radial_locality}

As seen in the previous sections, all the kernels separate into a part which depends only on the angular distance between pixels and another part that depends on the azimuthal orientation around the central pixel. The radial part of the kernels determines the non-locality of the operator. 
We numerically compute the radial kernels: $f, _{+2}f ~\&~ _{-2}f $ by evaluating the respective multipole sums given in \eq{eq:rad_ker_queb} and \eq{eq:f2_rad_ker} in the band limit $\ell \in [=2,192]$. The resultant functions are depicted in \fig{fig:beta_kernel}. 
%
\begin{figure}[!hbt]
\centering
\includegraphics[width=0.8\columnwidth]{kernel/beta_kernel.pdf}
\caption{The figure depicts the radial part of the convolution kernels. These radial function have been evaluated with the band limit fixed at $\ell \in [2,192]$. The vertical dashed line marks the approximate size of a NSIDE=64 Healpix pixel.}
\label{fig:beta_kernel}
\end{figure}
%
Note that the function $f(\beta)$, which forms the radial part of the real space convolution kernel that translates the Stokes parameters Q \& U to scalars E \& B, has a vanishing contribution from the location of the central pixel ($\beta \rightarrow 0$). Recall that the fields E \& B are scalar and hence expected to be immune to coordinate definitions. The locally defined Stokes parameters however necessarily depend on the coordinate definition. Therefore, this nature of the radial kernel is to be expected in order for it to satisfy the requirement of the derived quantities being coordinate independent.

The functions $_{+2}f(\beta)~\&~_{-2}f(\beta)$, both contribute to the convolution kernels which decompose the Stokes parameters into those corresponding to the respective scalar modes E \& B. $_{-2}f(\beta)$ form the radial part of the band limited delta function $I$ and hence it contributes the most at the location of the central pixel. $_{+2}f(\beta)$ again has a vanishing contribution near the central pixel and it dominantly contributes in the neighbouring regions which are approximately at least 1 pixel distance away from the central pixel.

It is clear from previous discussions that that the non-locality of the E and B modes is determined by the radial part of the convolution kernels. To quantify this non-locality as a function of the maximum multipole accessible for analysis, we evaluate the radial part of the convolution kernel for different values of $\ell_{\rm max}$, while keeping the lowest multipole fixed at $\ell_{\rm min}=2$. 
%
\begin{figure}[t]
\centering
\subfigure{\includegraphics[width=0.96\columnwidth]{kernel/f_rad_ker_fn_of_ellmax.pdf}}
\subfigure{\includegraphics[width=0.48\columnwidth]{kernel/fp2_rad_ker_fn_of_ellmax.pdf}}
\subfigure{\includegraphics[width=0.48\columnwidth]{kernel/fm2_rad_ker_fn_of_ellmax.pdf}}
\caption{The top panel shows a plot of the radial kernels $f(\beta,\ell_{\rm min},\ell_{\rm max})$ while the bottom left and right panels show the radial functions ${}_{+2}f(\beta,\ell_{\rm min},\ell_{\rm max}) ~\&~ {}_{-2}f(\beta,\ell_{\rm min},\ell_{\rm max})$ respectively, for different $\ell_{\rm max}$ as indicated by the legends and fixed $\ell_{\rm min}=2$. The curves for each of these functions have been normalized such that the maximum of the curve is set to unity. The horizontal solid black line marks the location where the amplitude of the kernel falls below 1\% of its maximum. The slanted dashed black lines indicate a power law fit (by eye) to the envelope of the radial functions. While the envelopes for function $f(\beta)~\&~ {}_{-2}f(\beta)$ are fit well by the power law $\propto \beta^{-1.5}$, the envelope for the function ${}_{-2}f(\beta)$ is seen to have a slightly steeper fall off $\propto \beta^{-1.65}$.}
\label{fig:rad_ker_decay}
\end{figure}
%
The set of radial kernels so derived are plotted in \fig{fig:rad_ker_decay}. All the function have been normalized such that their global maxima is set to unity. Note that on increasing $\ell_{\rm max}$ the radial kernels shift to the left, attaining fractions of  their global maxima at progressively small angular distances $\beta$ from the central pixel.

We have observed that the radial functions derived by evaluating the multipole sums to different maximum multipoles are self similar. Specifically these functions follow an interesting telescoping and scaling property: $${}_rf(\beta,2,\ell_{\rm max}) \approx \Big[\frac{\ell_{\rm max}}{\ell'_{\rm max}}\Big]^2{}_rf(\beta'=\frac{\ell_{\rm max}}{\ell'_{\rm max}} \beta ,2,\ell'_{\rm max}) \,,$$ where ${}_rf$ denotes all the different radial functions. We can understand the shifting left of the radial kernels on increasing the maximum multipole using this property. Lets say the function ${}_rf(\beta, \ell'_{\rm max})$ transition to being monotonously below some fraction of the global maxima at an angular distance of $\beta_0$.  The function ${}_rf(\beta', \ell_{\rm max})$, given $\ell_{\rm max} > \ell'_{\rm max}$, reaches the same transition point $\beta'=\beta_0$ at a smaller value of $\beta$ owing to the fact that $\ell_{\rm max}/\ell'_{\rm max}$ is greater than unity. The amplitude scaling of the functions is irrelevant since the transition point is always described in terms of the fraction of the global maxima of the function.

We are primarily interested in the non-local dependence of the scalar modes E \& B on the Stokes parameters Q \& U. We define the value of the abscissa at which the function $f(\beta,\ell_{\rm min},\ell_{\rm max})$ transits to being monotonously below 1\% of the maxima of the function as the non-locality parameter: $\beta_{o}$.  For a $\ell_{\rm max}=24$, the maximum multipole accessible on a Nside=8 Healpix map, the non-locality parameter $\beta_0=180^{\circ}$, since the radial function never falls monotonously below 1\% of the global maxima of the function. Using the self similar  property of the radial function discussed above, we define the following: $\beta_o= {\rm min}(180,180 \frac{24}{\ell_{\rm max}})$ as the non-locality parameter, given the maximum multipole $\ell_{\rm max}$. \revisit{$\beta_0$ is not very sensitive to $\ell_{\rm min}$, though it can be important on making drastic changes to $\ell_{\rm min}$. } \comment{Quantify how changing $\ell_{\rm min}$ changes $\beta_0$}

The envelope of the radial kernel $f(\beta,\ell_{\rm min},\ell_{\rm max})$ is observed to have a power law relation to the angular distance $\beta$ as seen in \fig{fig:rad_ker_decay}. At small values of $\beta$, which corresponds to the flat sky limit,  the radial function is proportional to $\beta^2$. In the intermediate range,  a fit by eye indicates that the envelope of the function is well represented by the power law $\beta^{-1.5}$.
\comment{There isn't too much of a discussion surrounding the function ${}_{\pm}f(\beta)$.}

>>>>

The real and imaginary part of the kernel $\mathcal{M}$ are identical irrespective of changes in the galactic latitude and longitude of the central pixel. Note that these functions are not distorted when a part of the domain overlaps with the poles, as can be seen in the first three rows of \fig{fig:vis_kernel}. Both these facts can be associated with the fact that this function does not depend on the Euler angle $\gamma$. From \eq{eq:qu2eb_convolution_explicit} and \eq{eq:eb2qu_convolution_explicit} it is clear that $\mathcal{M}_r$ and $\mathcal{M}_i$ can be interpreted in the following ways,
%\revisit{Recall that the angle $\gamma$ defines the rotation about the final z-axis which co-alligns the planar axes. Since the function $\mathcal{M}$ does not depend of this final orientation, it implies a rotational invariance of the fields constructed by convolving with this kernel.}
%
\begin{subequations}
\beqry
\bmat E= -\mathcal{M}_r \\ B =+\mathcal{M}_i  \emat  \leftarrow \bmat Q=\delta(\hat{n} - \hat{n}_j)\\ U=0 \emat ~~;~~ \bmat E= -\mathcal{M}_i \\ B = - \mathcal{M}_r  \emat  \leftarrow \bmat Q=0 \\ U=\delta(\hat{n} - \hat{n}_j) \emat \,,  \\
\bmat Q= -\mathcal{M}_r \\ U = -\mathcal{M}_i  \emat  \leftarrow \bmat E=\delta(\hat{n} - \hat{n}_j)\\ B=0 \emat ~~;~~ \bmat Q= +\mathcal{M}_i \\ U = - \mathcal{M}_r  \emat  \leftarrow \bmat E=0 \\ B=\delta(\hat{n} - \hat{n}_j) \emat \,.
\eeqry
\end{subequations}
%
The kernels $\mathcal{D}$ \& $\mathcal{I}$ vary significantly as a function of galactic latitude of the central pixel as seen in the last four columns of \fig{fig:vis_kernel}. These kernels show a two fold symmetry in the vicinity of the poles and this arises due to Euler angle $\gamma \approx 0$ here and therefore $e^{i2(\alpha \pm \gamma)} \approx e^{i2\alpha}$. Note that in this region, the azimuthal profile of the real and imaginary part of these kernels  is similar to $\mathcal{M}_r$ and $\mathcal{M}_i$ respectively.  This also explains why the imaginary part of the band limited delta function $\mathcal{I}$ contributes just as much as the real part in these regions. On moving to lower latitudes, $\mathcal{D}$ quickly transitions to having a four fold symmetry while $\mathcal{I}$ transitions to being dominated by the real part and behaves more like the conventional delta function. This transition can be most easily understood in the flat sky limit where $\gamma = -\alpha$ which leads to the resultant 4 fold symmetry seen for $\mathcal{D}$ owing to $e^{i2(\alpha - \gamma)} =e^{i4\alpha}$ and $\mathcal{I}$ being dominated by the real part owing to $e^{-i2(\alpha + \gamma)} =1 + i0$. Since the flat sky approximation has most validity in the proximity of the equator these limiting tendencies of the respective kernels are seen in the bottom row of \fig{fig:vis_kernel} which depict the kernels evaluated at the equator $b=0^{\circ}$. The middle two row depict the kernels evaluated at a latitudes of $b=87^{\circ}~\&~ 80^{\circ}$ and serve to indicate the rate of this transition. These kernels are invariant under changes in longitude of the central pixel, the latitude being held fixed, as one may have expected.

\subsection{Quantifying the non-locality of E \& B modes} \label{sec:radial_locality}
%
\begin{figure}[!hbt]
\centering
\includegraphics[width=0.8\columnwidth]{kernel/beta_kernel.pdf}
\caption{The figure depicts the radial part of the convolution kernels. These radial function have been evaluated with the band limit fixed at $\ell \in [2,192]$. The vertical dashed line marks the approximate Healpix pixel size of a NSIDE=64, which is the lowest resolution that allows access to $\ell_{\rm max}=192$.}
\label{fig:beta_kernel}
\end{figure}
%
\fig{fig:mixing_kernel} and the surrounding discussion provides a quantative understanding of the  azimuthal dependence of various kernels, however it is difficult to assess the radial nature of these kernels from these figures. The radial part determines the non-locality of the respective operators and encodes all the multipole dependence. We compute the radial kernels ${_{\mm}f}, {_{\md}f} ~\&~ {_{\mi}f} $ by evaluating the respective multipole sums in \eq{eq:rad_ker_queb} and \eq{eq:f2_rad_ker} in the band limit $\ell \in [2,192]$ and the resultant functions are depicted in \fig{fig:beta_kernel}.

Recall that ${_{\mm}f}$ is the radial part of the kernel that translates the Stokes parameters $Q$ \& $U$ to scalars $E$ \& $B$ and vice versa. Note that ${_{\mm}f}$ has a vanishing contribution from the location of the central pixel ($\beta \rightarrow 0$) as seen in \fig{fig:beta_kernel} and one can show that that ${_{\mm}f(\beta=\pi)}=0$. The coordinate dependence of the Stokes parameters cannot be integrated out in the vicinity of the locations $\beta=0,\pi$ due to the fact that the azimuthal angles become ill defined here therefore this nature of ${_{\mm}f}$ is necessary to ensure that the derived fields behave as scalars. Similarly while deriving the Stoke field from the scalars $E$ \& $B$ this nature of ${_{\mm}f}$ is necessary to ensure that the necessary coordinate dependencies are integrated in. ${_{\md}f}$ shows a similar behaviour, it has a vanishing contribution in the vicinity of the central pixel and dominantly contributes in regions which are approximately at least 1 pixel distance away from the central pixel as seen in \fig{fig:beta_kernel}. ${_{\mi}f}$ is the radial part of the band limited delta function $\mi$ and expectedly contributes the most at the location of the central pixel.

\noindent \textit{The band limit dependence:} It is clear from previous discussions that the scalar field $E$ \& $B$ constructed at a location depends on the Stokes field in the surrounding regions. We further quantify this non-locality by studying the radial extent of the kernels and its dependence on the maximum multipole accessible for analysis. To carry out this study we evaluate the radial functions for different values of $\ell_{\rm max}$, while keeping the lowest multipole fixed at $\ell_{\rm min}=2$. 
%
\begin{figure}[t]
\centering
\subfigure{\includegraphics[width=0.96\columnwidth]{kernel/f_rad_ker_fn_of_ellmax.pdf}}
\subfigure{\includegraphics[width=0.48\columnwidth]{kernel/fp2_rad_ker_fn_of_ellmax.pdf}}
\subfigure{\includegraphics[width=0.48\columnwidth]{kernel/fm2_rad_ker_fn_of_ellmax.pdf}}
\caption{The top panel depicts the radial function  ${_{\mm}f}(\beta,\ell_{\rm min},\ell_{\rm max})$ while the bottom left and right panels show the radial functions ${}_{\md}f(\beta,\ell_{\rm min},\ell_{\rm max}) ~\&~ {}_{\mi}f(\beta,\ell_{\rm min},\ell_{\rm max})$ respectively, for different $\ell_{\rm max}$ as indicated by their legends and fixed $\ell_{\rm min}=2$. All the curves have been normalized such that the maximum of the curve is set to unity. The horizontal solid black line marks the location where the amplitude of the kernel falls below 1\% of its maximum. The slanted dashed black lines indicate a power law fit (by eye) to the envelope of the radial functions. While the envelopes for function ${_{\mm}f}(\beta)~\&~ {}_{\mi}f(\beta)$ are fit well by the power law $\propto \beta^{-1.5}$, the envelope for the function ${}_{\md}f(\beta)$ is seen to have a slightly steeper slope $\propto \beta^{-1.65}$.}
\label{fig:rad_ker_decay}
\end{figure}
%
The resultant set of radial function are depicted in \fig{fig:rad_ker_decay}, where all the function have been normalized such that their global maxima is set to unity. We note that on increasing $\ell_{\rm max}$ the radial kernels shift left, attaining their global maxima at progressively small angular distance from the central pixel. The amplitude of these radial function scales up as $\propto \ell_{\rm max}^2$. At intermediate values of $\beta$, the envelope of the radial functions is fit well by a power law $ \propto \beta^{-n}$, the details of this fit can be seen in \fig{fig:rad_ker_decay}.
We observe that the radial functions computed by evaluating the multipole sums to different maximum multipoles are self similar and follow an interesting telescoping and scaling property, $${}_rf(\beta,2,\ell_{\rm max}) \approx \Big[\frac{\ell_{\rm max}}{\ell'_{\rm max}}\Big]^2{}_rf(\beta'=\frac{\ell_{\rm max}}{\ell'_{\rm max}} \beta ,2,\ell'_{\rm max}) \,,$$ where ${}_rf$ denotes all the different radial functions. 
%We can understand the shifting left of the radial kernels on increasing the maximum multipole using this property. Lets say the function ${}_rf(\beta, \ell'_{\rm max})$ transition to being monotonously below some fraction of the global maxima at an angular distance of $\beta_0$.  The function ${}_rf(\beta', \ell_{\rm max})$, given $\ell_{\rm max} > \ell'_{\rm max}$, reaches the same transition point $\beta'=\beta_0$ at a smaller value of $\beta$ owing to the fact that $\ell_{\rm max}/\ell'_{\rm max}$ is greater than unity. The amplitude scaling of the functions is irrelevant since the transition point is always described in terms of the fraction of the global maxima of the function.

It is useful to define a characteristic radius of the region from which the scalar fields evaluated at a point get most of their contribution from.  Since the primarily interest is in the non-locality of the scalar modes $E$ \& $B$ we define the abscissa at which the function ${_{\mm}f}(\beta,\ell_{\rm min}=2,\ell_{\rm max})$ transits to being monotonously below 1\% of the maxima of the function as the non-locality parameter: $\beta_{o}$.  For $\ell_{\rm max}=24$, the maximum multipole accessible on a Nside=8 Healpix map, the non-locality parameter $\beta_0=180^{\circ}$ as the radial function never falls monotonously below 1\% of its global maxima. Using this fact and the self similar property of the radial functions, we define the following empirical relation: $\beta_o= {\rm min}(180,180 \frac{24}{\ell_{\rm max}})$, as a means of estimating the non-locality parameter given the maximum multipole $\ell_{\rm max}$ accessible for analysis.
